---
title: "Major League Baseball Payroll and Wins Analysis"
author: "Maria Elizabeth Abad"
date: "February 19, 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(kableExtra)
library(cowplot)
```

# 1. Wrangle

## 1.1 Import the Data
```{r import-data}
load("C:/Users/Elize/Documents/YR3 2nd sem/DATA MINING/ml_pay.rdata")
ls()

mlb_raw <- as_tibble(ml_pay)
load("C:/Users/Elize/Documents/YR3 2nd sem/DATA MINING/ml_pay.rdata")
ls()

mlb_raw <- as_tibble(ml_pay)  # Adjust object name if needed
head(mlb_raw)

cat("Rows:", nrow(mlb_raw), "Columns:", ncol(mlb_raw))
```

**Answer:** This matches up with the data description given.

## 1.2 Tidying the Data (with pct_wins calculation)
```{r tidy-data}
mlb_yearly <- mlb_raw %>%
  pivot_longer(cols = -c(Team.name.2014, payroll, avgwin),
               names_to = "column_name",
               values_to = "value") %>%
  separate(column_name, into = c("prefix", "year", "suffix"), sep = "(?<=\\D)(?=\\d)", fill = "right") %>%
  mutate(tidy_col_name = case_when(
    prefix == "p" ~ "payroll",         
    TRUE ~ "num_wins"                 
  )) %>%
  select(Team.name.2014, year, tidy_col_name, value) %>%
  pivot_wider(names_from = tidy_col_name, values_from = value) %>%
  rename(team = Team.name.2014) %>%
  mutate(pct_wins = (num_wins / 162) * 100)

head(mlb_yearly)
```

**Answer:** `mlb_aggregate` contains 30 rows (one per team). `mlb_yearly` contains 510 rows (17 years for each team).

## 1.3 Calculate Aggregate Data
```{r create-mlb-aggregate}
mlb_aggregate <- mlb_yearly %>%
  group_by(team) %>%
  summarise(
    payroll_aggregate = sum(payroll, na.rm = TRUE),
    pct_wins_aggregate = mean(pct_wins, na.rm = TRUE)
  )

head(mlb_aggregate)
```

## Quality Control
```{r quality-control}
mlb_aggregate_computed <- mlb_yearly %>%
  group_by(team) %>%
  summarise(
    payroll_aggregate_computed = sum(payroll, na.rm = TRUE),
    pct_wins_aggregate_computed = mean(pct_wins, na.rm = TRUE)
  )

mlb_aggregate_joined <- mlb_aggregate %>%
  left_join(mlb_aggregate_computed, by = "team")

print(mlb_aggregate_joined)

p1 <- ggplot(mlb_aggregate_joined, aes(x = payroll_aggregate, y = payroll_aggregate_computed)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  ggtitle("Payroll Aggregate: Provided vs. Computed") +
  xlab("Provided Payroll Aggregate") +
  ylab("Computed Payroll Aggregate")

p2 <- ggplot(mlb_aggregate_joined, aes(x = pct_wins_aggregate, y = pct_wins_aggregate_computed)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  ggtitle("Win Percentage: Provided vs. Computed") +
  xlab("Provided Win Percentage Aggregate") +
  ylab("Computed Win Percentage Aggregate")

library(cowplot)
plot_grid(p1, p2)
```

# 2. Explore

## 2.1 Payroll Across Years
```{r payroll-across-years}
mean_payrolls <- mlb_yearly %>%
  group_by(team) %>%
  summarise(mean_payroll = mean(payroll, na.rm = TRUE))

ggplot(mlb_yearly, aes(x = as.numeric(year), y = payroll)) +
  geom_line() +
  facet_wrap(~ team) +
  geom_hline(data = mean_payrolls, aes(yintercept = mean_payroll), color = "red", linetype = "dashed") +
  ggtitle("Payroll Across Years by Team") +
  xlab("Year") +
  ylab("Payroll (in millions)")

# Identify the three teams with the greatest payroll_aggregate_computed
top_payroll_teams <- mlb_aggregate_computed %>%
  arrange(desc(payroll_aggregate_computed)) %>%
  head(3)

print(top_payroll_teams)

payroll_wide <- mlb_yearly %>%
  select(team, year, payroll) %>%
  pivot_wider(names_from = year, values_from = payroll, names_prefix = "payroll_")

payroll_increase <- payroll_wide %>%
  mutate(pct_increase = ((payroll_2014 - payroll_1998) / payroll_1998) * 100) %>%
  arrange(desc(pct_increase)) %>%
  select(team, payroll_1998, payroll_2014, pct_increase)

top_pct_increase_teams <- payroll_increase %>% head(3)
print(top_pct_increase_teams)
```

## 2.2 Win Percentage Across Years
```{r win-percentage-years}
mean_pct_wins <- mlb_yearly %>%
  group_by(team) %>%
  summarise(mean_pct_wins = mean(pct_wins, na.rm = TRUE))

ggplot(mlb_yearly, aes(x = as.numeric(year), y = pct_wins)) +
  geom_line() +
  facet_wrap(~ team) +
  geom_hline(data = mean_pct_wins, aes(yintercept = mean_pct_wins), color = "red", linetype = "dashed") +
  ggtitle("Win Percentage Across Years by Team") +
  xlab("Year") +
  ylab("Win Percentage (%)")

top_pct_wins_teams <- mlb_aggregate_computed %>%
  arrange(desc(pct_wins_aggregate_computed)) %>%
  head(3)

print(top_pct_wins_teams)

pct_wins_sd <- mlb_yearly %>%
  group_by(team) %>%
  summarise(pct_wins_sd = sd(pct_wins, na.rm = TRUE)) %>%
  arrange(desc(pct_wins_sd))

top_erratic_teams <- pct_wins_sd %>% head(3)
print(top_erratic_teams)
```

## 2.3 Win Percentage versus Payroll
```{r win-percentage-payroll}
ggplot(mlb_aggregate_computed, aes(x = payroll_aggregate_computed, y = pct_wins_aggregate_computed)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_text_repel(aes(label = team), size = 3) +
  ggtitle("Win Percentage vs. Payroll (Aggregated)") +
  xlab("Payroll Aggregate (in millions)") +
  ylab("Win Percentage Aggregate (%)")
```

## 2.4 Team Efficiency
```{r team-efficiency}
mlb_aggregate_computed <- mlb_aggregate_computed %>%
  mutate(efficiency = pct_wins_aggregate_computed / payroll_aggregate_computed)

top_efficient_teams <- mlb_aggregate_computed %>%
  arrange(desc(efficiency)) %>%
  select(team, efficiency, pct_wins_aggregate_computed, payroll_aggregate_computed) %>%
  head(3)

print(top_efficient_teams)
```
